---
title: "Calculating Movement Path Metrics in R"
output: html_document
author: "Briana Abrahms"
date: "September 10th 2019"
---
```{r libraries, message=FALSE}
library(tidyverse)
library(sf) #for simple feature spatial data
library(move) #for loading data from Movebank and calculating some path metrics
library(adehabitatLT) # for calculating large suite of path metrics
library(knitr)
```

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, cache = TRUE)
knitr::opts_knit$set(root.dir = '~/Dropbox/Research/R Tutorial Scripts & Workshops/Abrahms UW Movement Analysis Workshop/Workshop Materials/')
```

# Movement data
Movement data may come in a variety of scales and resolutions, 
requiring careful consideration before analysis. 

Here we are going to focus on GPS relocation data as it is one of the most common
forms of tracking data and provides a (mostly) consistent record of an animals movement path. 

There are many packages in R used to handle movement data, and
an ever growing number built for specific analyses. Unfortunately there is no strict
consensus yet on which class of movement data objects should be in across packages, and researchers often 
have to be comfortable converting from one to the other depending on the task at hand and the 
format required by each package. Although there are definitely more R packages
created for dealing with animal movement data, I am going to focus today on the two most
prominent: `move` and `adehabitatLT`.

To begin we are going to introduce Movebank and `Move` & `MoveStack` objects with the 
library `move`. We will then introduce the `adehabitatLT` package and demonstrate
how easy it is to pull out primary stepwise characteristics like step length, turning
angle, and net displacement from our trajectories. These primary path characteristics
often form the basis for many more complex movement analyses. 

## Movebank & the `move` package
Though often researchers in your position
will have data of your own, there has been an effort 
to share movement data from other collections through the Movebank database. Movebank 
is a free online infrastructure created to help researchers manage, share, analyze, 
and archive animal movement data. `move` is an R package that contains functions to
access movement data stored in [movebank.org](http://www.movebank.org) as well
as tools to visualize and statistically analyze animal movement data. 
`move` is addressing movement ecology questions that complement existing efforts 
such as `adehabitat` and other packages which we will introduce later.

There are 3 ways to load movement data into R as a `move` object:

1. Using the movebank API
2. Directly loading movebank-formatted .csvs
3. Converting data from a traditional dataframe 

Here let's consider some Albatross data, from movebank study #2911040. To load 
data directly from the movebank API like I will here, you would need your own 
movebank username and password, here I
have stored my password in my local environment as the variable `pass`. 

```{r direct Movebank download, eval=FALSE}
loginStored <- movebankLogin(username="briana.abrahms", password=local(pass, env=local.env))

#let's look at a study with albatross data
getMovebankStudy(2911040, login=loginStored)$citation

#if you want to search for studies by a search term
# searchMovebankStudies(x="albatross", login=loginStored)

#check for animal IDs - 
getMovebankAnimals(2911040, login=loginStored)
#get the all data - getMovebankData()
#get only specific animal by specifying individual_id

albatross <- getMovebank("event", study_id = 2911040, login=loginStored)

head(albatross)

# readr::write_tsv(albatross, "Study2911040") #save as write.csv but faster and without row names
# getMovebank("event", login, study_id,...): returns the sensor measurements from a study. 
# See also getMovebankData, getMovebankNonLocationData.
```

To save us the effort of getting you all Movebank logins, I have provided
the tsv. We can use this to demonstrate the 3rd way of loading in movement data.
```{r convert dataframe to move, message=FALSE}
albatross <- read_tsv("data_files/Study2911040", comment="##", col_types = "Tnncc") %>% # Tnncc = time, number, number, character, character
  as.data.frame() %>%
  na.omit()

#create 'move' object                       
albatross_move <- move(x=albatross$location_long,
              y=albatross$location_lat,
              time=albatross$timestamp,
              data=albatross,
              animal = as.factor(albatross$individual_id),
              sensor = as.factor(albatross$tag_id),
              proj=CRS("+proj=longlat +ellps=WGS84"))

albatross_move
head(albatross_move)
```
We've created our own MoveStack, which is a stack of dataframes for each individual. We can plot it using the move::plot command. 

```{r}
plot(albatross_move, type="l") # all of them
```

If we are curious to compare behaviors across individuals, we can split the moveStack, 
into separte move objects with the command `split`
```{r}
ids <- split(albatross_move)
```

This allows us to plot and manipulate the individuals separately:
```{r id move plot}
par(mfrow=c(2,2))
plot(ids[[1]], type='l', main= names(ids[1]))
plot(ids[[2]], type='l', main= names(ids[2]))
plot(ids[[3]], type='l', main= names(ids[3]))
plot(ids[[4]], type='l', main= names(ids[4]))
```

```{r echo=FALSE}
par(mfrow=c(1,1))
```


We can manipulate, and clean up, this data easily when converted to an sf object from the sf ("simple features") package for working with simple spatial data. 

### Side note: the `sf` package

The **sf** library is an R implementation of:  
 - a new spatial data class system in R  
 - functions for reading and writing spatial data  
 - tools for spatial operations on vectors  
 
Ultimately this seeks to replace the older **sp**, **rgdal**, **rgeos** packages 
which formed the original toolset for working with spatial data in R. The **sf** library
replaces the S4 class structure used in **sp** with simple feature access - the current standard
across industry for organizing spatial data -- extending R's data.frame structure directly
to accept spatial geometry attributes and making it easier to manipulate spatial datasets
using tools like dplyr and the tidyverse. However, as this package is new and under 
developement there are times were we will switch back to the S4 class structure to play
nice with our movement packages. 

More information regarding this shift [here](https://www.r-consortium.org/blog/2017/01/03/simple-features-now-on-cran)

```{r make sf}
albatross_sf <- 
  albatross %>% 
  st_as_sf(coords = c("location_long", "location_lat"), crs=4326)
```

This comes in handy for some of the data visualization we will be doing later. In addition, in this state, it's simple to interogate
the data, such as looking at how many records we have per individual:
```{r tally}
albatross_sf %>% 
  group_by(individual_id) %>% 
  tally()

# we could do something similar on the movestack by calling...
table(albatross_move@trackId)
```

## Calculate Path Characteristics
We can use additional functions in the `move` package to calculate some path characteristics.

- Step Length: the distance between successive relocations is often used in animal
movement analysis (e.g. Root and Kareiva 1984, Marsh and Jones 1988). The `move` package
automatically calculates distances in meters if the coordinates are projected in long/lat 
degrees, so no conversion is necessary.
```{r step length move}
steplengths <- distance(albatross_move)

hist(steplengths[[1]])
```

- Turning Angle
Your turning angle measures the change of direction between
the step built by relocations i − 1 and i and the step built by relocations
i and i + 1 (what we generally refer to as “turning angle”). It is often used together with
the parameter dist to fit movement models (e.g. Root and Kareiva 1984,
Marsh and Jones 1988). 
```{r turn angles move}
turnangles <- turnAngleGc(albatross_move)

hist(turnangles[[1]])
```

- Speed: the speed between successive relocations (in meters/second for long/lat projection) can often help identify if there are
any outliers based on unrealistic movement speeds.
```{r speeds move}
speeds <- speed(albatross_move)

hist(speeds[[1]])
```

There is lots more the `move` package has to offer, but we are going to move
onto ltraj objects and the primary path characteristics
they make so easy to extract. 

## AdehabitatLT package

The 'adehabitatLT' package works with 'ltraj' objects. The class ltraj is intended to store trajectories of animals. `ltraj` objects function neatly as lists, similar in some ways 
to the list of Move objects that was created when we split out albatross MoveStack. 
A key difference between Move and ltraj objects is not only in the structure of data
but also the attributes calculated. Ltraj ojects automatically calculate common path
characteristics like step length, relative turning angle, absolute turning angle, and
net squared displacement from which much of subsequent movement modelling builds off of. 
Additionally from these primary path characteristics it's simple to calculate 
secondary statistics like sinuosity, residence time, or directional persistence. 

`ltrajs` can be created from a traditional data frames, spatial dataframes, or 
directly from move objects using the command `as.ltraj` and, in the case of the 
dataframes, specifying the coordinates. 

Before we create an ltraj object below, we will first add in UTM coordinates. This is 
necessary for calculating path metrics in meters because unlike the `move` package,
`adehabitatLT` does not automatically convert long/lat projections into meters.

```{r ltraj object, message=FALSE}
#add UTM coordinates
coords <- sp::SpatialPoints(data.frame(albatross$location_long, albatross$location_lat), proj4string=CRS("+proj=longlat +ellps=WGS84")) 
UTM_coords <- spTransform(coords, CRS("+proj=utm +south +zone=16 +ellps=WGS84")) # reproject to UTM (meters)
albatross$utm_long <- UTM_coords@coords[,1]
albatross$utm_lat <- UTM_coords@coords[,2]

# create 'ltraj' object
alba_ltraj  <- as.ltraj(xy=cbind(albatross$utm_long, albatross$utm_lat),
                        date=albatross$timestamp,
                        id=as.factor(albatross$individual_id),
                        proj4string = CRS("+proj=utm +south +zone=16 +ellps=WGS84"))

alba_ltraj  

# note that we can also convert our move object directly into an ltraj object, but that will retain the long/lat projection
# alba_ltraj  <- as(albatross_move, 'ltraj')
```

Note the information the this class gives us up front. Not only can we see the 
number of relocations from each of our animals we can see exactly the start and end dates 
of each trajectory.

We can quickly plot all individuals or just specific ones.
```{r ltraj plot}
plot(alba_ltraj)

plot(alba_ltraj[1:3])
```


# Automatic Path Characteristics
If we dig into any one of our ltraj list objects, we'll see a dataframe 
containing certain basic characteristics from our paths. Specifically we get 
**step length**, **turn angle**, and **net squared displacement** automatically
whenever we create an ltraj object. That is very convenient! Keep in mind
the x and y coordinates need to be in UTM in order for distance and displacement
metrics to be calculated in meters.

```{r ltraj dataframe}
#let's look at a dataframe in the ltraj list
#it is very easy in the ltraj list format to grab different individuals
#for example here is the dataframe for the first individual

head(alba_ltraj[[1]])
```

Here we see step length (`dist`), turn angle (`rel.angle`), absolute angle (`abs.angle`), and
net squared displacement (`R2n`). Whereas the relative turning angle measures the change of direction between
the step built by relocations i − 1 and i and the step built by relocations
i and i + 1, the absolute angle is the angle between the x direction and the step
built by relocations i and i + 1. 
```{r angles}
hist(alba_ltraj[[1]]$rel.angle)

hist(alba_ltraj[[1]]$abs.angle)
```

- Net Squared Displacement, i.e R2n,the squared distance between the first relocation of the trajectory
and the current relocation is very significant in animal movement ecology. It is often used to test some movements models
(e.g. the correlated random walk, see the seminal paper of Kareiva and
Shigesada, 1983). It is also widely used to identify broad-scale movement patterns
like range residency, migration, nomadism, dispersal (see Bunnefeld et al. 2010).

```{r NSD}
hist(alba_ltraj[[1]]$R2n)

#let's look at NSD over time
plot(alba_ltraj[[1]]$date, alba_ltraj[[1]]$R2n, xlab="date", ylab="NSD") #this looks like classic migratory behavior
```

Note the distributions plotted above as they often form the basis for accurate simulations
of animal movement paths. 

From these metrics we can also calculate secondary characteristics such as *velocity*,
*straightness*, *sinuosity*, *directional persistence*,  *persistence velocity*, *residence time*, 
*time to return*, and more.

In addition to the derivation of these primary and secondary path characteristics, 
adehabitatLT is especially useful for path segementation and some simple movement models. 
If interested, I highly recommend consulting the package's very thourough documentation. 
You can also explore vignettes of the adehabitatLT package [here](https://cran.r-project.org/web/packages/adehabitatLT/).

### Side note: types of trajectories in adehabitatLT

The adehabitat packages, and ltraj objects, distinguish between 2 types of trajectories:
1. Trajectories of type I are characterized by the fact that the time is not
precisely known or not taken into account for the relocations of the
trajectory (i.e. sampling of tracks in snow)
2. Trajectories of type II are characterized by the fact that the time is
known for each relocation. This type of trajectory is in turn be divided
into two subtypes:
  – regular trajectories: these trajectories are characterized by a constant
time lag between successive relocations;
  – irregular trajectories: these trajectories are characterized by a
variable time lag between successive relocations

It's worth emphasizing that functions in `adehabitatLT` are mainly designed to deal with
type I or type II regular trajectories. Irregular trajectories are harder to analyze as their 
descriptive parameters may not be compared when computed on different time lags.

